# (PART) All is fail {-}

# Local and immediate debugging

R scripts are great things, when they work. What are some strategies you can
use when things go sideways and an error occurs in your script?

## Debugging your own code

The most common case you will run into a bug is when writing new code yourself.
Often the mistake is obvious and easily fixed, but sometimes it only appears
after multiple levels of calls and is harder to diagnose. There are a few
common strategies to use when debugging your own code.

- Use `traceback()` to determine where a given error is occurring.
- Output diagnostic information in code with `print()`, `cat()` or `message()` statements.
- Use `browser()` to open an interactive debugger before the error
- Use `debug()` to automatically open a debugger at the start of a function call.
- Use `trace()` to start a debugger at a location inside a function.

### traceback()

The `traceback()` function can be used to print a summary of how your program
arrived at the error. This is also called a call stack, stack trace or
backtrace.

In R this gives you each call that lead up to the error, which can be very
useful for determining what lead to the error.

You can use `traceback()` in two different ways, either by calling it
immediately after the error has occurred.

<!-- knitr messes with the tracebacks, so we just do this manually :| -->

```{r, eval = FALSE}
f <- function(x) x + 1
g <- function(x) f(x)
g("a")
```
```
#> Error in x + 1 : non-numeric argument to binary operator
```
```r
traceback()
```
```
#> 2: f(x) at #1
#> 1: g("a")
```

Or by using `traceback()` as an error handler, which will call it immediately
on any error. (You could even put this in your [`.Rprofile`](#rprofile))

```{r, eval = FALSE}
options(error = traceback)
g("a")
```

```
#> Error in x + 1 : non-numeric argument to binary operator
#> 2: f(x) at #1
#> 1: g("a")
```

### `print()`

Once you know where an error occurs it is then helpful to know why. Often
errors occur because functions are given inputs their authors did not expect,
so it is useful to print the value of objects during execution.

The most basic way to do this is to sprinkle messages throughout your code,
with `print()` or `str()`. `str()` is often more useful because it gives more detail
into the exact structure of an object, which may not be the structure you
expect it to be.

The main downsides to the print approach is you often have to add them in
multiple places to narrow down the error, and you cannot further investigate
the object.

### `browser()`

A more sophisticated debugging method is to put a call to `browser()` in your
code. This will stop execution at that point and open R's interactive debugger.
In the debugger you can run any R command to look at objects in the current
environment, modify them and continue executing.

Some useful things to do are

1. Use `ls()` to determine what objects are available in the current
   environment. This allows you to see exactly what things you can examine.
2. Use `str()`, `print()` etc. to examine the objects
3. Use `n` to evaluate the next statement. Use `s` to evaluate the next
   statement, but step into function calls.
4. Use `where` to print a [stack trace](#traceback)
5. Use `c` to leave the debugger and continue execution
6. Use `Q` to exit the debugger and return to the R prompt.

TODO: RStudio debugging

### `debug()`

If you have control of the code (because you are the one writing it), using
`browser()` is generally the most convenient way to enter the debugger. However
if the error is occurring in code in a package what options do you have?

This is where the `debug()` function is useful, it will open the R debugger on
any function, including those in packages.

TODO: good example / activity

```r
debug(ggplot2::ggplot)
```

Use can use the `::` syntax to find 'exported' functions in a package, but
there is also a way to access _any_ function, including un-exported ones,
`:::`.

```r
debug(ggplot2:::set_last_plot)
```

### `trace()`

`debug()` is very useful, but one drawback is it always executes the first time
a function is called. What can you do if the bug only happens the 100th time a
function is called?

`trace()` is a more flexible version of `debug()` that not only lets you start
a debugger at the start of a function, it lets you insert _any_ code at _any_
location in a function. The downside to this power and flexibility is that
`trace()` is comparatively harder to use than `debug()`.

### `recover()`

`recover()` is generally not used directly, instead it is used as an error
handler, by calling `options(error = recover)`. The benefit to `recover()` over
using `options(error = browser)` is that you can browse on any of the call
stack, not just where the error occurred. Often the issue is above The benefit
to `recover()` over using `options(error = browser)` is that you can browse on
any of the call stack, not just where the error occurred. Often the issue is
most easily diagnosed in calls higher on the call stack than immediately where
the error occurred.

When recover is called it prints a list of the current calls, with a prompt to
select which you want to browse in. Then [`browser()`](#browser) is invoked at
that location.

### Resources

- [Debugging techniques in RStudio - Amanda Gadrow](https://resources.rstudio.com/wistia-rstudio-conf-2018-2/debugging-techniques-in-rstudio-amanda-gadrow-4)

## TODO
The first thing you should do is try and to read and understand the _whole_
error message. If this does not provide much information a good first step
would be to google the error message or do a search for it on
<https://community.rstudio.com> or <https://stackoverflow.com>.


Gabe once shared a method of debugging Rmd: https://gist.github.com/gmbecker/e47b0951b91505b2f4607259bcec8f97. Interesting example of `options(error = recover)`.
